Northern Pikeminnow Predation on Juvenile Chinook Salmon and Potential Impacts to Local Populations in the Upper Salmon River, Idaho

### 2021-12-20 analysis/paper/paper.Rmd
minor edits to abundance methods

We estimated the population size of piscine predators in the Deadwater Slough using a mark-recapture survey and a catch per unit effort (CPUE) approach. Predators were sampled near the peaks of the juvenile DSR and NRR emigrations. Our intent was to sampling during the emigration of the 2018 brood year (fall 2019 and spring 2020). However, due to logistical concerns during the onset of the Covid-19 pandemic, the spring 2020 effort was delayed until fall 2020. Although this provided the benefit of an additional abundance estimate during the DSR emigration, estimates during the NRR emigration were still lacking, and so an additional effort was added in spring 2021. Fall efforts included sampling across a two-week span; however, due to budget constraints the spring 2021 effort was conducted over a single week. Sampling was constrained within a two-week window to minimize predator emigration/immigration and to constrain the population for the mark-recapture models.

@@

Multiple capture methods were employed during the fall 2019 effort to reduce selectivity and bias for species and size classes. Methods included raft electrofishing, fyke netting, snorkeling, and angling. However, ineffective methods were abandoned in subsequent years. After evaluating all methods, angling proved to be the most effective method for capturing piscine predators while minimizing potential impacts to ESA-listed steelhead, but recognizing that angling can have both species and size selectivity (bias). From here on, we focus on fish captured by angling, unless otherwise noted. Additionally, it quickly became clear during fall 2019 that Northern Pikeminnow were by far the most prevalent piscine predator in Deadwater Slough and so is the focus of our study.

<!-- For example, fish were able to elude the electrical field of the raft electrofisher due to 1) the lack of flow in Deadwater Slough and 2) our inability to "chase" fish downstream as they evaded the electricity. Snorkeling was ineffective due to the high turbidity of the mainstem Salmon River and Deadwater Slough area. Significant amounts of debris in the water column also created concerns that fyke nets could become entangled. Additionally, there were concerns that raft electrofishers and fyke nets could incidentally capture ESA-listed adult steelhead that are present in the mainstem Salmon River during fall and spring. -->

@@

Mark-recapture abundance estimators fell into two broad categories: single-census and multiple-census. For the single-census estimators, we used the Lincoln-Peterson estimator:

$$
\hat{N} = \frac{(M)(n)}{(m)}
$$

where $M$ is the number is the total number of fish marked and returned to the population during the first week of sampling (mark event), $n$ is the total number of fish caught in the second week (recapture event), and $m$ is the number of marked fish caught during the recapture event.

Because the Lincoln-Peterson estimator can be biased with small samples, we also employed the Chapman-modified Lincoln-Peterson estimator:

$$
\hat{N} = \frac{(M + 1)(n + 1)}{(m + 1)} - 1
$$

For a multiple-census estimator, we first utilized the Schnabel estimator:

@@

where $M$, $n$, and $m$ are indexed by the sampling occassion, $i$. Here, $i$ is defined as each sampling day. The Schnabel estimator does not have an associated standard error; however, the 90% confidence intervals were calculated.

Finally, we explored one additional multiple-census estimator, the Schumacher-Eschmeyer estimator, which is based on minimizing the weighted sum of squares between the proportion of marked individuals in the sample and the unknown proportion of marked individuals in the population. The Schumacher-Eschmeyer estimator is as follows:

@@

The mark-recapture estimators allowed us to estimate the abundance of Northern Pikeminnow during the fall 2019 and fall 2020 sampling periods. However, due to logistical constraints, we were only able to sample during a single week in spring 20201 and released fish were left unmarked. Instead, to estimate the abundance of Northern Pikeminnow during spring 2021, we used the ratio of total CPUE in the fall sampling events to the total CPUE in the spring sampling event, and multiplied that ratio by the average abundance of Northern Pikeminnow from the fall sampling events.<!-- Kevin, is this previous statement accurate? --> This approach assumes equal capture probabilities between the fall and spring capture events. If that assumption is not true and capture probability did differ, we believe that the capture probability  in the spring would be lower due to higher flows, which would lead to our spring abundance estimate to be biased low. Therefore, we presume our abundance estimate of Northern Pikeminnow for the spring 2021 sampling event to be a conservative estimate.

>>>>

We estimated the population size of piscine predators in the Deadwater Slough using a mark-recapture survey and a catch per unit effort (CPUE) approach. Predators were sampled near the peaks of the juvenile DSR and NRR emigrations. Our intent was to sample during the emigration of the 2018 brood year (fall 2019 and spring 2020). However, due to logistical concerns during the onset of the Covid-19 pandemic, the spring 2020 effort was delayed until fall 2020. Although this provided the benefit of an additional abundance estimate during the DSR emigration, estimates during the NRR emigration were still lacking, and so an additional effort was added in spring 2021. Fall efforts included sampling across a two-week span; however, due to budget constraints the spring 2021 effort was conducted over a single week. Fall efforts were constrained within a two-week window to minimize predator emigration/immigration and to constrain the population for the mark-recapture models.

@@

Multiple capture methods were employed during the fall 2019 effort to reduce selectivity and bias for species and size classes. Methods included raft electrofishing, fyke netting, snorkeling, and angling. However, ineffective methods were abandoned in subsequent years. <!-- For example, fish were able to elude the electrical field of the raft electrofisher due to 1) the lack of flow in Deadwater Slough and 2) our inability to "chase" fish downstream as they evaded the electricity. Snorkeling was ineffective due to the high turbidity of the mainstem Salmon River and Deadwater Slough area. Significant amounts of debris in the water column also created concerns that fyke nets could become entangled. Additionally, there were concerns that raft electrofishers and fyke nets could incidentally capture ESA-listed adult steelhead that are present in the mainstem Salmon River during fall and spring. -->After evaluating all methods, angling proved to be the most effective method for capturing piscine predators while minimizing potential impacts to ESA-listed steelhead. <!--, but recognizing that angling can have both species and size selectivity (bias).--> Therefore, the following analyses will focus on fish captured by angling, unless otherwise noted. During the fall 2019 effort it additionally became clear that Northern Pikeminnow were by far the most prevalent piscine predator in Deadwater Slough and so is the focus of our study.

@@

Four different mark-recapture abundance estimators were employed to explore variation in estimates. Estimators fell into two broad categories: single-census and multiple-census. For the single-census estimators, we used the Lincoln-Peterson estimator:

$$
\hat{N} = \frac{(M)(n)}{(m)}
$$
where $M$ is the number is the total number of fish marked and returned to the population during the first week of sampling (mark event), $n$ is the total number of fish caught in the second week (recapture event), and $m$ is the number of marked fish caught during the recapture event. Because the Lincoln-Peterson estimator can be biased with small samples, we also employed the Chapman-modified Lincoln-Peterson estimator:

$$
\hat{N} = \frac{(M + 1)(n + 1)}{(m + 1)} - 1
$$
For a multiple-census estimator, we first utilized the Schnabel estimator:

@@

where $M$, $n$, and $m$ are indexed by the sampling occassion, $i$. Here, $i$ is defined as each sampling day. The Schnabel estimator does not have an associated standard error; however, the 90% confidence intervals were calculated. Lastly, we explored one additional multiple-census estimator, the Schumacher-Eschmeyer, which is based on minimizing the weighted sum of squares between the proportion of marked individuals in the sample and the unknown proportion of marked individuals in the population. The Schumacher-Eschmeyer estimator is as follows:

@@

The mark-recapture estimators allowed us to estimate the abundance of Northern Pikeminnow during the fall 2019 and fall 2020 sampling periods. However, to estimate the abundance of Northern Pikeminnow during spring 2021, which was not a mark-recapture design, we used the ratio of total CPUE in the fall sampling events to the total CPUE in the spring sampling event, then multiplied that ratio by the average abundance from the fall sampling events.<!-- Kevin, is this previous statement accurate? --> This approach assumes equal capture probabilities between the fall and spring capture events. <!--If that assumption is not true and capture probability did differ, we believe that the capture probability  in the spring would be lower due to higher flows, which would lead to our spring abundance estimate to be biased low. Therefore, we presume our abundance estimate of Northern Pikeminnow for the spring 2021 sampling event to be a conservative estimate.-->
